Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves

### 2025-07-21 ms/manuscraps.qmd
updated pai analysis

>>>>

#### Light-response (\aqcurve) curves

In `r percent_aa_and_aq`\% of plants, we measured light-response (\aqcurve) curves on the same leaflets as \agcurve{} curves. However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for \aqcurve{} curves. Leaves acclimated to high light-intensity (\ppfdequals{2000}), ambient CO$_2$ (\caequals{415}), \rhequals{50}, and \tleafequals{25}. After $A$ and \gsw{} stabilized, we measured $A$ at 20 light-intensity levels between $0$ and $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ in descending order.

### 2025-07-21 ms/ms.qmd
updated pai analysis

#### Light-response (\aqcurve) curves

In `r percent_aa_and_aq`\% of plants, we measured light-response (\aqcurve) curves on the same leaflets as \agcurve{} curves. However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for \aqcurve{} curves. Leaves acclimated to high light-intensity (\ppfdequals{2000}), ambient CO$_2$ (\caequals{415}), \rhequals{50}, and \tleafequals{25}. After $A$ and \gsw{} stabilized, we measured $A$ at 20 light-intensity levels between $0$ and $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ in descending order.

@@

$$\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx$$
where $\theta_\text{amphi}$ and $\theta_\text{hypo}$ are the quadratic parameters of the amphi and pseudohypo \agcurve{} curves, respectively (@fig-example-aa). We calculated $\widehat{\mathrm{AA}}$ from each draw of the posterior for the paired amphi and pseudohypo \agcurve{} curves to obtain the posterior distribution of \aax{} for each leaf at each light intensity. We then summarized the posterior distribution of \aax{} for each leaf at each light intensity by the posterior median and standard deviation, which became our point estimate of, and uncertainty in, \aax{} for the phylogenetic mixed effects model described below.

```{r, example-aa, echo=FALSE,message=FALSE,warning=FALSE}
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 3.5
#| fig-cap: "Example amphistomy advantage (\\aax) calculation from \\agcurve{} curves in an amphistomatous leaf with gas exchange through both surfaces (blue) and the same leaf with gas exchange blocked on the upper surface only (orange). We integrate over the gray region between the curves to estimate \\aax{}. The \\aax{} is low (left facet) when the photosynthetic rate of an amphistomatous leaf is similar to a pseudohypostomatous leaf at the same total stomatal conductance ($x$-axes); large \\aax{} (right facet) is indicated when an amphistomatous leaf has a higher photosynthetic rate than a pseudohypostomatous leaf."
read_rds("../figures/example-aa.rds")
```

@@

We fit Bayesian mixed effects models with phylogenetically structured random effects to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats. We fit all models combinations of measurement light intensity, growth light intensity, and their interaction as both fixed effects and random effects among populations. In other words, we test whether there were main effects of light intensity treatments and/or whether populations varied in their response to light intensity treatments. In all models, we included population as a phylogenetically structured random effect. We accounted for uncertainty in estimates of \aax{} using the standard deviation of the posterior distribution of each \aax{} estimate (@sec-aa-est). We crossed fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}. In total, we fit `r n_model1` models, each with a different combination of fixed, random, and distributional effects. We used the RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic windows from @pease_phylogenomics_2016. Two of our populations were not in this tree. We used accession LA1044 in place of LA3909, two populations of *S. galapagnse*; LA0750 was added as sister to LA0716, two closely related populations of *S. pennellii*. The node separating LA0750 from LA0716 was placed half-way between the next deepest node. We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-models}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values. We selected the model with the lowest LOOIC value to generate posterior predictions for hypothesis testing.

@@

## MARKER - need to update table

```{r, predictions, eval=TRUE, echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-predictions
#| tbl-cap: "Predictions of competing hypotheses about the relationship between \\aax{} and light intensity, light treatment, and native \\ppfd{} among populations. The middle column lists specific directional predictions about parameter values and model fit according to leave-one-out cross-validation information criterion (LOOIC), where $\\text{LOOIC}_i$ is the LOOIC value for model $i$. The rightmost column describes the predictions in words and explains how population-level \\aax{} is calculated in the relevant model."

tribble(
  ~"\\textbf{Hypothesis}", ~"\\textbf{Prediction(s)}", ~"\\textbf{Description}",
  "Assimilatory", "$\\beta_{\\mathrm{AA},2000} > 0$", "Average \\aax{} at high light intensity is greater than that at low light intensity",
  "Plastic", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "Average \\aax{} in sun leaves is greater than that in shade leaves",
  "Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} ($\\mathrm{AA}_\\text{acc}$) increases with native \\ppfd",
  "Constitutive", "", "$\\mathrm{AA}_\\text{acc} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}}$",
  "Assimilatory $\\times$ Plastic", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "Average \\aax{} is highest at high light intensity in sun leaves",
  "Assimilatory $\\times$ Plastic", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{2}$", "",
  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000} > 0$", "Average \\aax{} at high light intensity is greater than that at low light intensity",
  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{acc},2000}$) increases with native \\ppfd",
  "Assimilatory $\\times$ Constitutive", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{3}$", "$\\mathrm{AA}_{\\text{acc},2000} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}}$",
  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "Average \\aax{} in sun leaves is greater than that in shade leaves",
  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{acc},\\text{sun}}$) increases with native \\ppfd",
  "Plastic $\\times$ Constitutive", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{4}$", "$\\mathrm{AA}_{\\text{acc},\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$",
  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "Average \\aax{} is highest at high light intensity in sun leaves",
  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} at high light intensity in sun leaves ($\\mathrm{AA}_{\\text{acc},2000,\\text{sun}}$) increases with native \\ppfd",
  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\text{LOOIC}_\\text{3} > \\text{LOOIC}_\\text{5}$", "$\\mathrm{AA}_{\\text{acc},2000,\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$"

>>>>

#### Light-saturated photosynthetic rate

In `r percent_aa_and_aq`\% of plants, we measured the maximum photosynthetic rate ($A_\mathrm{max}$) on the same leaflets as \agcurve{} curves. However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for $A_\mathrm{max}$. Leaves acclimated to high light-intensity (\ppfdequals{2000}), ambient CO$_2$ (\caequals{415}), \rhequals{50}, and \tleafequals{25}. After $A$ and \gsw{} stabilized, we measured $A$ at $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$.

@@

$$\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx$$
where $\theta_\text{amphi}$ and $\theta_\text{hypo}$ are the quadratic parameters of the amphi and pseudohypo \agcurve{} curves, respectively (@fig-example-aa). We calculated $\widehat{\mathrm{AA}}$ from each draw of the posterior for the paired amphi and pseudohypo \agcurve{} curves to obtain the posterior distribution of $\widehat{\mathrm{AA}}$ for each leaf at each light intensity. We then summarized the posterior distribution of $\widehat{\mathrm{AA}}$ for each leaf at each light intensity by the posterior median and standard deviation, which became our point estimate of, and uncertainty in, $\widehat{\mathrm{AA}}$ for the phylogenetic mixed effects model described below.

```{r, example-aa, echo=FALSE,message=FALSE,warning=FALSE}
#| label: fig-example-aa
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 3.5
#| fig-cap: "Example amphistomy advantage (\\aax) calculation from \\agcurve{} curves in an amphistomatous leaf with gas exchange through both surfaces (blue) and the same leaf with gas exchange blocked on the upper surface only (orange). We integrate over the gray region between the curves to estimate \\aax{}. The \\aax{} is low (left facet) when the photosynthetic rate of an amphistomatous leaf is similar to a pseudohypostomatous leaf at the same total stomatal conductance ($x$-axes); large \\aax{} (right facet) is indicated when an amphistomatous leaf has a higher photosynthetic rate than a pseudohypostomatous leaf."
read_rds("../figures/example-aa.rds")
```

@@

We fit Bayesian mixed effects models with phylogenetically structured random effects to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats. We fit all models combinations of measurement light intensity, growth light intensity, and their interaction as both fixed effects and random effects among populations. In other words, we test whether there were main effects of light intensity treatments and/or whether populations varied in their response to light intensity treatments. In all models, we included population as a phylogenetically structured random effect. We accounted for uncertainty in $\widehat{\mathrm{AA}}$ using the standard deviation of the posterior distribution of each $\widehat{\mathrm{AA}}$ estimate (@sec-aa-est). We crossed fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}. In total, we fit `r n_model1` models, each with a different combination of fixed, random, and distributional effects. We used the RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic windows from @pease_phylogenomics_2016. Two of our populations were not in this tree. We used accession LA1044 in place of LA3909, two populations of *S. galapagnse*; LA0750 was added as sister to LA0716, two closely related populations of *S. pennellii*. The node separating LA0750 from LA0716 was placed half-way between the next deepest node. We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-aa_loo1}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values. We selected the model with the lowest LOOIC value to generate posterior predictions for hypothesis testing.

@@

```{r, predictions, eval=TRUE, echo=FALSE,message=FALSE,warning=FALSE}
#| label: tbl-predictions
#| tbl-cap: "Predictions of competing hypotheses about the relationship between \\aax{} and light intensity, light treatment, and native plant area index (PAI) among populations. The middle column lists specific directional predictions about parameter values. The rightmost column describes the predictions in words and explains how population-level \\aax{} is calculated in the relevant model."

tribble(
  ~"\\textbf{Hypothesis}", ~"\\textbf{Prediction(s)}", ~"\\textbf{Description}",
  "Assimilatory", "$\\beta_{\\mathrm{AA},2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
  "Plastic", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
  "Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} ($\\mathrm{AA}_\\text{acc}$) decreases with native PAI",
  "Constitutive", "", "$\\mathrm{AA}_\\text{acc} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}}$",
  "Assimilatory $\\times$ Plastic", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "\\aax{} is highest at high light intensity in sun leaves",
  "Assimilatory $\\times$ Plastic", "", "",
  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{acc},2000}$) decreases with native PAI",
  "Assimilatory $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{acc},2000} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}}$",
  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{acc},\\text{sun}}$) decreases with native PAI",
  "Plastic $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{acc},\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$",
  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "\\aax{} is highest at high light intensity in sun leaves",
  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} at high light intensity in sun leaves ($\\mathrm{AA}_{\\text{acc},2000,\\text{sun}}$) decreases with native PAI",
  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{acc},2000,\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$"

### 2025-07-21 ms/ms.tex
updated pai analysis

\paragraph{\texorpdfstring{Light-response (\aqcurve)
curves}{Light-response () curves}}\label{light-response-curves}

In 91.3\% of plants, we measured light-response (\aqcurve) curves on the
same leaflets as \agcurve{} curves. However, when a leaflet was damaged
during \agcurve{} curves, we used the next closest leaflet for
\aqcurve{} curves. Leaves acclimated to high light-intensity
(\ppfdequals{2000}), ambient CO\(_2\) (\caequals{415}), \rhequals{50},
and \tleafequals{25}. After \(A\) and \gsw{} stabilized, we measured
\(A\) at 20 light-intensity levels between \(0\) and
\(\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\) in
descending order.

@@

\[\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx\]
where \(\theta_\text{amphi}\) and \(\theta_\text{hypo}\) are the
quadratic parameters of the amphi and pseudohypo \agcurve{} curves,
respectively (\textbf{?@fig-example-aa}). We calculated
\(\widehat{\mathrm{AA}}\) from each draw of the posterior for the paired
amphi and pseudohypo \agcurve{} curves to obtain the posterior
distribution of \aax{} for each leaf at each light intensity. We then
summarized the posterior distribution of \aax{} for each leaf at each
light intensity by the posterior median and standard deviation, which
became our point estimate of, and uncertainty in, \aax{} for the
phylogenetic mixed effects model described below.

\begin{figure}[H]

{\centering \pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/example-aa-1.pdf}}

@@

\caption{Example amphistomy advantage (\aax) calculation from \agcurve{}
curves in an amphistomatous leaf with gas exchange through both surfaces
(blue) and the same leaf with gas exchange blocked on the upper surface
only (orange). We integrate over the gray region between the curves to
estimate \aax{}. The \aax{} is low (left facet) when the photosynthetic
rate of an amphistomatous leaf is similar to a pseudohypostomatous leaf
at the same total stomatal conductance (\(x\)-axes); large \aax{} (right
facet) is indicated when an amphistomatous leaf has a higher
photosynthetic rate than a pseudohypostomatous leaf.}

@@

We fit Bayesian mixed effects models with phylogenetically structured
random effects to test predictions of competing hypotheses about why
amphistomy advantage (\aax) might be greater for leaves in sunny, open
habitats. We fit all models combinations of measurement light intensity,
growth light intensity, and their interaction as both fixed effects and
random effects among populations. In other words, we test whether there
were main effects of light intensity treatments and/or whether
populations varied in their response to light intensity treatments. In
all models, we included population as a phylogenetically structured
random effect. We accounted for uncertainty in estimates of \aax{} using
the standard deviation of the posterior distribution of each \aax{}
estimate (Section~\ref{sec-aa-est}). We crossed fixed and random effect
structures with fixed effects of measurement light intensity and/or
growth light intensity on the distributional parameter, \(\sigma\), the
residual variance in \aax{}. In total, we fit 100 models, each with a
different combination of fixed, random, and distributional effects. We
used the RAxML whole-transcriptome concatenated phylogeny based on 2,745
100-kb genomic windows from (\emph{64}). Two of our populations were not
in this tree. We used accession LA1044 in place of LA3909, two
populations of \emph{S. galapagnse}; LA0750 was added as sister to
LA0716, two closely related populations of \emph{S. pennellii}. The node
separating LA0750 from LA0716 was placed half-way between the next
deepest node. We used the leave-one-out cross-validation information
criterion (LOOIC) to compare the fit of models (\autoref{tbl-models})
using the \emph{R} package \textbf{loo} version 2.8.0 (\emph{65}) to
calculate LOOIC values. We selected the model with the lowest LOOIC
value to generate posterior predictions for hypothesis testing.

@@

\caption{\label{tbl-predictions}Predictions of competing hypotheses
about the relationship between \aax{} and light intensity, light
treatment, and native \ppfd{} among populations. The middle column lists
specific directional predictions about parameter values and model fit
according to leave-one-out cross-validation information criterion
(LOOIC), where \(\text{LOOIC}_i\) is the LOOIC value for model \(i\).
The rightmost column describes the predictions in words and explains how
population-level \aax{} is calculated in the relevant model.}

@@

\toprule
\textbf{Hypothesis} & \textbf{Prediction(s)} & \textbf{Description}\\
\midrule
Assimilatory & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}Average \aax{} at high light intensity is greater than that at low light intensity\\
\cmidrule{1-3}\pagebreak[0]
Plastic & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}Average \aax{} in sun leaves is greater than that in shade leaves\\
\cmidrule{1-3}\pagebreak[0]
Constitutive & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} ($\mathrm{AA}_\text{acc}$) increases with native \ppfd\\
\nopagebreak
 &  & \hspace{-1em}$\mathrm{AA}_\text{acc} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}}$\\
\cmidrule{1-3}\pagebreak[0]
Assimilatory $\times$ Plastic & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}Average \aax{} is highest at high light intensity in sun leaves\\
\nopagebreak
 & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{2}$ & \hspace{-1em}\\
\cmidrule{1-3}\pagebreak[0]
Assimilatory $\times$ Constitutive & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}Average \aax{} at high light intensity is greater than that at low light intensity\\
\nopagebreak
 & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity ($\mathrm{AA}_{\text{acc},2000}$) increases with native \ppfd\\
\nopagebreak
 & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{3}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}}$\\
\cmidrule{1-3}\pagebreak[0]
Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}Average \aax{} in sun leaves is greater than that in shade leaves\\
\nopagebreak
 & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} in sun leaves ($\mathrm{AA}_{\text{acc},\text{sun}}$) increases with native \ppfd\\
\nopagebreak
 & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{4}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
\cmidrule{1-3}\pagebreak[0]
Assimilatory $\times$ Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}Average \aax{} is highest at high light intensity in sun leaves\\
\nopagebreak
 & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity in sun leaves ($\mathrm{AA}_{\text{acc},2000,\text{sun}}$) increases with native \ppfd\\
\nopagebreak
 & $\text{LOOIC}_\text{3} > \text{LOOIC}_\text{5}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000,\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
\bottomrule

>>>>

\paragraph{Light-saturated photosynthetic
rate}\label{light-saturated-photosynthetic-rate}

In 91.3\% of plants, we measured the maximum photosynthetic rate
(\(A_\mathrm{max}\)) on the same leaflets as \agcurve{} curves. However,
when a leaflet was damaged during \agcurve{} curves, we used the next
closest leaflet for \(A_\mathrm{max}\). Leaves acclimated to high
light-intensity (\ppfdequals{2000}), ambient CO\(_2\) (\caequals{415}),
\rhequals{50}, and \tleafequals{25}. After \(A\) and \gsw{} stabilized,
we measured \(A\) at
\(\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\).

@@

\[\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx\]
where \(\theta_\text{amphi}\) and \(\theta_\text{hypo}\) are the
quadratic parameters of the amphi and pseudohypo \agcurve{} curves,
respectively (Figure~\ref{fig-example-aa}). We calculated
\(\widehat{\mathrm{AA}}\) from each draw of the posterior for the paired
amphi and pseudohypo \agcurve{} curves to obtain the posterior
distribution of \(\widehat{\mathrm{AA}}\) for each leaf at each light
intensity. We then summarized the posterior distribution of
\(\widehat{\mathrm{AA}}\) for each leaf at each light intensity by the
posterior median and standard deviation, which became our point estimate
of, and uncertainty in, \(\widehat{\mathrm{AA}}\) for the phylogenetic
mixed effects model described below.

\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-example-aa-1.pdf}}

@@

\caption{\label{fig-example-aa}Example amphistomy advantage (\aax)
calculation from \agcurve{} curves in an amphistomatous leaf with gas
exchange through both surfaces (blue) and the same leaf with gas
exchange blocked on the upper surface only (orange). We integrate over
the gray region between the curves to estimate \aax{}. The \aax{} is low
(left facet) when the photosynthetic rate of an amphistomatous leaf is
similar to a pseudohypostomatous leaf at the same total stomatal
conductance (\(x\)-axes); large \aax{} (right facet) is indicated when
an amphistomatous leaf has a higher photosynthetic rate than a
pseudohypostomatous leaf.}

@@

We fit Bayesian mixed effects models with phylogenetically structured
random effects to test predictions of competing hypotheses about why
amphistomy advantage (\aax) might be greater for leaves in sunny, open
habitats. We fit all models combinations of measurement light intensity,
growth light intensity, and their interaction as both fixed effects and
random effects among populations. In other words, we test whether there
were main effects of light intensity treatments and/or whether
populations varied in their response to light intensity treatments. In
all models, we included population as a phylogenetically structured
random effect. We accounted for uncertainty in \(\widehat{\mathrm{AA}}\)
using the standard deviation of the posterior distribution of each
\(\widehat{\mathrm{AA}}\) estimate (Section~\ref{sec-aa-est}). We
crossed fixed and random effect structures with fixed effects of
measurement light intensity and/or growth light intensity on the
distributional parameter, \(\sigma\), the residual variance in \aax{}.
In total, we fit 100 models, each with a different combination of fixed,
random, and distributional effects. We used the RAxML
whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic
windows from (\emph{64}). Two of our populations were not in this tree.
We used accession LA1044 in place of LA3909, two populations of \emph{S.
galapagnse}; LA0750 was added as sister to LA0716, two closely related
populations of \emph{S. pennellii}. The node separating LA0750 from
LA0716 was placed half-way between the next deepest node. We used the
leave-one-out cross-validation information criterion (LOOIC) to compare
the fit of models (\autoref{tbl-aa_loo1}) using the \emph{R} package
\textbf{loo} version 2.8.0 (\emph{65}) to calculate LOOIC values. We
selected the model with the lowest LOOIC value to generate posterior
predictions for hypothesis testing.

@@

\caption{\label{tbl-predictions}Predictions of competing hypotheses
about the relationship between \aax{} and light intensity, light
treatment, and native plant area index (PAI) among populations. The
middle column lists specific directional predictions about parameter
values. The rightmost column describes the predictions in words and
explains how population-level \aax{} is calculated in the relevant
model.}

@@

\toprule
\textbf{Hypothesis} & \textbf{Prediction(s)} & \textbf{Description}\\
\midrule
Assimilatory & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}\aax{} at high light intensity is greater than that at low light intensity\\
\cmidrule{1-3}\pagebreak[0]
Plastic & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}\aax{} in sun leaves is greater than that in shade leaves\\
\cmidrule{1-3}\pagebreak[0]
Constitutive & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} ($\mathrm{AA}_\text{acc}$) decreases with native PAI\\
\nopagebreak
 &  & \hspace{-1em}$\mathrm{AA}_\text{acc} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}}$\\
\cmidrule{1-3}\pagebreak[0]
Assimilatory $\times$ Plastic & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}\aax{} is highest at high light intensity in sun leaves\\
\nopagebreak
 &  & \hspace{-1em}\\
\cmidrule{1-3}\pagebreak[0]
Assimilatory $\times$ Constitutive & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}\aax{} at high light intensity is greater than that at low light intensity\\
\nopagebreak
 & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity ($\mathrm{AA}_{\text{acc},2000}$) decreases with native PAI\\
\nopagebreak
 &  & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}}$\\
\cmidrule{1-3}\pagebreak[0]
Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}\aax{} in sun leaves is greater than that in shade leaves\\
\nopagebreak
 & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} in sun leaves ($\mathrm{AA}_{\text{acc},\text{sun}}$) decreases with native PAI\\
\nopagebreak
 &  & \hspace{-1em}$\mathrm{AA}_{\text{acc},\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
\cmidrule{1-3}\pagebreak[0]
Assimilatory $\times$ Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}\aax{} is highest at high light intensity in sun leaves\\
\nopagebreak
 & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity in sun leaves ($\mathrm{AA}_{\text{acc},2000,\text{sun}}$) decreases with native PAI\\
\nopagebreak
 &  & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000,\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
\bottomrule
